---
title: "Week3 Assignment"
author: "Amol Suraj Mishra"
date: "17 October 2017"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Instructions

Create a web page presentation using R Markdown that features a plot created with Plotly. Host your webpage on either GitHub Pages, RPubs, or NeoCities. 

Your webpage must contain the date that you created the document, and it must contain a plot created with Plotly. We would love to see you show off your creativity!

## Project Summary

This assignment is to display some information using plotly presentation.

```{r echo=FALSE, warning=FALSE, message=FALSE }
library(ggplot2)
library(plotly)
library(gapminder)
```

## Slide with Plot

```{r echo=FALSE, warning=FALSE, message=FALSE }
gDat <- gapminder
jDat <- droplevels(subset(gDat, continent != "Oceania"))

# colorsURL <- "http://www.stat.ubc.ca/~jenny/notOcto/STAT545A/examples/gapminder/data/gapminderCountryColors.txt"
# countryColors <- read.delim(file = colorsURL, as.is = 3) # protect color

countryColors <- read.delim(file="CountryColors", header=TRUE, as.is = 3)

jColors <- countryColors$color
names(jColors) <- countryColors$country

jYear <- 2007

jDat <- jDat[with(jDat, order(year, -1 * pop)), ]
gap.bubble.plot <- ggplot(subset(jDat, year == jYear),
       aes(x = gdpPercap, y = lifeExp )) + scale_x_log10() +
  geom_point(aes(size = sqrt(pop/pi), text = paste(country, lifeExp, "years")), pch = 21) + 
  scale_size_continuous(range=c(1,40)) +
  facet_wrap(~ continent) + 
  aes(fill = country) + scale_fill_manual(values = jColors) + theme(legend.position="none") +
  xlab("GDP Per Capita\n") + ylab("\nLife Expectancy")


ggplotly(gap.bubble.plot, tooltip=c("text"))

```

